Translation of B Implementations to the LLVM: Control and Modularity

### 2013-09-05 translation.tex
Completed a new version of the translation definition updating support for "IMPORT" clauses and "CALL" operations. Updated Python module bimp.py to build abstract syntax tree for B implementations. Registered a new example of B implementation with a single imported machine, not prefixed. Created the Python data structure corresponding to that example. obs. 1: The implementation of the translation for import clauses is not completed. obs. 2: The construction Python data structures of existing examples needs to be synchronized with respect to bimp.py.

        \B{op} & : & \B{Name} \\
        \B{pre} & : & opt \B{Name} \\

@@

        \B{out} & : & seq \B{Name}

@@

\DD{Specify \B{Skip}.}

@@

The translation unit is the B implementation, say \B{M}, and is mapped to a LLVM
module, say \llvm{M}. A B implementation \B{M} has several sections, each
translated as follows:
\begin{enumerate}
\item Visibility clauses provide access to definitions found in external
  B components and are translated to external declarations in LLVM.
\item Constants and value definitions are translated to their value in the
  LLVM platform.
\item To encode the implementation state, the translation produces a
  structure type, named \llvm{\%M\$state\$}, and an instance of that type, named
  \llvm{@M\$self\$}. All the variables and instances of imported machines
  are encoded as elements of that structure.  Also, for each operation of each
  imported machine, an external function declaration is produced.
\item Each operation \B{op} of the implementation is translated into the
  definition of a function named \llvm{@M\$op}; also the translation produces a
  special function \llvm{@M\$init\$} responsible for the initialization.

@@

In those rules, $\PH{t}$ indicates a text place-holder where the value of $t$
shall be inserted and $\nl$ denotes a new line. For instance, we define
$\local(\B{n}) \isdef \llvm{"\%\PH{\B{n.id}}"}$ as a function that given an
element \B{n} of the abstract syntax tree returns the corresponding LLVM local
name, obtained by prepending character \% to its identifier; and $\Global(\B{n})
\isdef \llvm{"@\PH{\B{n.root.id}}\$\PH{\B{n.id}}"}$ yields the corresponding
LLVM global name, composed by special characters (to avoid name conflicts), the
name of the component where it is defined and its own name.  Also, we define
three auxiliary functions responsible to produce such identifiers of,
respectively, the type representing the state of a B module \B{n}, its
initialization function, and the object carrying its state:
\begin{align*}
\state(\B{n}) & \isdef \llvm{"\%\PH{\B{n.id}}\$state\$"} & \text{ (the type for the state)} \\
\init(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$init\$"} & \text{ (the initialization function)} \\
\self(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$self\$"} & \text{ (the instance in a top-level module)}
\end{align*}
We assume an unlimited supply of labels and local variables: we use $\lbl$ to
return a fresh LLVM label and $\variable$ for a fresh local variable name.

@@

\begin{align*}
\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L \isdef & \IF \B{il} \mbox{ is empty} \THEN \llvm{"br label \%\PH{\ell} \nl"} \\
& \ELSE \\
& \LET \B{i}, \B{il'} = \B{il}  \IN \\
& \quad \quad \LET p_1 = 
\begin{array}[t]{l}
  \IF \B{i} \mbox{ is \B{If} or \B{While}} \THEN \\
  \quad \LET \ell' = \lbl \IN \llvm{"\PH{\trad{i, $\ell'$}{\B{Inst}}_L} \PH{\ell'}: \PH{p}"} \\
  \ELSE \llvm{"\PH{\trad{i}{\B{Inst}}} \PH{p}"} \IN
\end{array} \\
& \quad \quad \IN \\
& \quad \quad \quad \LET p_2 = \trad{il', $\ell$}{\ListOf{\B{Inst}}}_L \\
& \quad \quad \quad \quad \llvm{"\PH{p_1} \PH{p_2}"}
\end{align*}

@@

@@

\paragraph{Call up} instructions apply B operations with some values and obtain
results. The LLVM code generated must include instructions to evaluate the
operation arguments, the address of the variables receiving the results as well
as an instance of the LLVM \llvm{call} instruction. In addition, the proposed
translation includes as argument the address of a structure storing the state of
the component corresponding to this operation. Note that this component may be
either the implementation being translated, or an instance of an imported
machine.
\begin{align*}
  \trad{c}{\B{Call}} \isdef 
  & \LET p_i \sep il = \trad{c.inp}{\ListOf{Inp}} \AND p_o \sep ol = 
  \trad{c.out}{\ListOf{Out}} \AND \\
  & \quad id = \IF \B{c.pre} = \text{None} \THEN \Global(\B{c}) \ELSE \llvm{"@\PH{c.pre}\$\PH{c.id}"} \AND \\
  & \quad ty = \IF \B{c.pre} = \text{None} \THEN \state(\B{c.root}) \ELSE \llvm{"\%\PH{c.pre}\$state\$"} \AND \\
  & \quad val = \IF \B{c.pre} = \text{None} \THEN \llvm{"\%self\$"} \ELSE \\
  & \quad \quad \llvm{"getelementptr \PH{\state(\B{n.root})} \%self\$, i32 0, i32 \PH{\idx(\B{c.pre})}"}\\
  & \IN \\
  & \quad \llvm{"\PH{p_i}} \\
  & \quad \llvm{\PH{p_o}} \\
  & \quad \llvm{call void \PH{id}(\PH{ty}* \PH{val}, \PH{il}, \PH{ol}) \nl"}
\end{align*}
The translation of input and output arguments produces two code snippets. The
first corresponds to the LLVM instruction sequence $p$ responsible for
evaluating the argument, and the second to the type $t$ and name $v$ of the
variable storing the resulting value:
\begin{align*}
  \trad{i}{Inp} \isdef 
  & \LET p \sep v \sep t = \trad{i}{\B{Expr}} \IN p \sep \llvm{"\PH{t} \PH{v}"}\\
  \trad{o}{Out} \isdef 
  & \LET p \sep v \sep t = \trad{o}{\B{Expr}} \IN p \sep \llvm{"\PH{t}* \PH{v}"}
\end{align*}

>>>>

        \B{op} & : & \B{Oper} \\

@@

        \B{out} & : & seq \B{Name} \\
        \B{inst} & : & opt \B{Impo}

@@

@@

The translation unit is the B implementation, say \B{M}, and is mapped to a LLVM
module, say \llvm{M}. A B implementation \B{M} has several sections, each
translated as follows:
\begin{enumerate}
\item Visibility clauses provide access to definitions found in external
  B components and are translated to external declarations in LLVM.
\item Constants and value definitions are translated to their value in the
  LLVM platform.
\item To encode the implementation state, the translation produces a structure
  type, named \llvm{\%M\$state\$}, and an instance of that type, named
  \llvm{@M\$self\$}. All the variables and instances of imported machines are
  encoded as elements of that structure. These elements are referenced by their
  position in the structure within the LLVM code. Also, for each operation of
  each imported machine, an external function declaration is produced.
\item Each operation \B{op} of the implementation is translated into the
  definition of a function named \llvm{@M\$op}; also the translation produces a
  special function \llvm{@M\$init\$} responsible for the initialization.

@@

In those rules, $\PH{t}$ indicates a text place-holder where the value of $t$
shall be inserted and $\nl$ denotes a new line. For instance, we define
$\local(\B{n}) \isdef \llvm{"\%\PH{\B{n.id}}"}$ as a function that given an
element \B{n} of the abstract syntax tree returns the corresponding LLVM local
name, obtained by prepending character \% to its identifier; and $\Global(\B{n})
\isdef \llvm{"@\PH{\B{n.root.id}}\$\PH{\B{n.id}}"}$ yields the corresponding
LLVM global name, composed by special characters (to avoid name conflicts), the
name of the component where it is defined and its own name.  Also, we define
three auxiliary functions responsible to produce such identifiers:
\begin{align*}
\state(\B{n}) & \isdef \llvm{"\%\PH{\B{n.id}}\$state\$"} & \text{ (the type for the state)} \\
\init(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$init\$"} & \text{ (the initialization function)} \\
\self(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$self\$"} & \text{ (the instance in a top-level module)}
\end{align*}
We assume an unlimited supply of labels and local variables: we use $\lbl$ to
return a fresh LLVM label and $\variable$ for a fresh local variable name.

Finally, we need a function that, given a concrete variable, or an imported
machine, returns the position of the corresponding element in the structure
representing the state of the implementation. The position of the elements in
the structure is the same as the position of corresponding B construct in the
source code:
\begin{align*}
  \idx(\B{n}) & \isdef \mathit{index}(\B{n}, \B{n.root.impo} \conc
\B{n.root.vari})
\end{align*}
where $\mathit{index}(e, s)$ is a function that returns the position of $e$ in
sequence $s$ and $\conc$ is the concatenation operator for sequences.

@@

\begin{align*}
\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L \isdef & \IF \B{il} \mbox{ is empty} \THEN \llvm{"br label \%\PH{\ell} \nl"} \\
& \ELSE \\
& \LET \B{i}, \B{il'} = \B{il}  \IN \\
& \quad \quad \LET p_1 = 
\begin{array}[t]{l}
  \IF \B{i} \mbox{ is \B{If} or \B{While}} \THEN \\
  \quad \LET \ell' = \lbl \IN \llvm{"\PH{\trad{i, $\ell'$}{\B{Inst}}_L} \PH{\ell'}: \PH{p}"} \\
  \ELSE \llvm{"\PH{\trad{i}{\B{Inst}}} \PH{p}"} \IN
\end{array} \\
& \quad \quad \IN \\
& \quad \quad \quad \llvm{"\PH{p_1}} \\
& \quad \quad \quad \llvm{\PH{\trad{il', $\ell$}{\ListOf{\B{Inst}}}_L}"}
\end{align*}

@@

\paragraph{Call up} instructions apply B operations with some values and obtain
results. The LLVM code generator produces instructions to evaluate the operation
arguments, the address of the variables receiving the results as well as an
instance of the LLVM \llvm{call} instruction. In addition, the proposed
translation includes as argument the address of a structure storing the state of
the component corresponding to this operation. Note that this component may be
either the implementation being translated (in the case of a so-called local
operation), or an instance of an imported machine.
\begin{align*}
  \trad{c}{\B{Call}} \isdef 
  & \LET p_i \sep il = \trad{c.inp}{\ListOf{Inp}} \AND p_o \sep ol = 
  \trad{c.out}{\ListOf{Out}} \AND \\
  & \quad id = \Global(\B{c}) \AND ty = \state(\B{c.root}) \AND \\
  & \quad val = \IF \B{c.inst} = \text{None} \THEN \llvm{"\%self\$"} \ELSE \\
  & \quad \quad \llvm{"getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{c.inst})}"}\\
  & \IN \\
  & \quad \llvm{"\PH{p_i}} \\
  & \quad \llvm{\PH{p_o}} \\
  & \quad \llvm{call void \PH{id}(\PH{ty}* \PH{val}, \PH{il}, \PH{ol}) \nl"}
\end{align*}
The translation of input and output arguments produces two code snippets. The
first corresponds to the LLVM instruction sequence $p$ responsible for
evaluating the argument, and the second to the type $t$ and name $v$ of the
variable storing the resulting value:
\begin{align*}
  \trad{i}{Inp} \isdef 
  & \LET p \sep v \sep t = \trad{i}{\B{Expr}} \IN p \sep \llvm{"\PH{t} \PH{v}"}\\
  \trad{o}{Out} \isdef 
  & \LET p \sep v \sep t = \trad{o}{\B{Expr}} \IN p \sep \llvm{"\PH{t}* \PH{v}"}
\end{align*}

\paragraph{Identity} does nothing and no code is generated:
\begin{align*}
\begin{split}
  \trad{n}{\B{Skip}} \isdef & \llvm{""}
\end{split}
\end{align*}

@@
